\documentclass[11pt]{article}

\author{wilricknl}
\title{3D Math Primer - Solutions chapter 1}

\usepackage{amsmath}
\usepackage{enumerate}

\begin{document}

\maketitle

\section{Solutions chapter 2}

\subsection{Exercise 1}

\begin{enumerate}[a.]
    \item
    \begin{itemize}
        \item \textbf{a} is a two dimensional row vector.
        \item \textbf{b} is a three dimensional column vector.
        \item \textbf{c} is a four dimensional column vector.
    \end{itemize}
    \item $0 + 6 - 3 + 5 = 8$
\end{enumerate}

\subsection{Exercise 2}

\begin{enumerate}[a.]
	\item Scalar.
	\item Scalar.
	\item Vector: direction north and magnitude 2 blocks.
	\item 600 mph is scalar and altitude is also scalar.
\end{enumerate}


\subsection{Exercise 3}

\begin{enumerate}[a.]
	\item % a
$
\begin{bmatrix}
0 & 2
\end{bmatrix}
$
	\item % b
$
\begin{bmatrix}
0 & -2
\end{bmatrix}
$
	\item % c
$
\begin{bmatrix}
0.5 & 2
\end{bmatrix}
$
	\item % d
$
\begin{bmatrix}
0.5 & 2
\end{bmatrix}
$
	\item % e
$
\begin{bmatrix}
0.5 & -3
\end{bmatrix}
$
	\item % f
$
\begin{bmatrix}
-2 & 0
\end{bmatrix}
$
	\item % g
$
\begin{bmatrix}
-2 & 1
\end{bmatrix}
$
	\item % h
$
\begin{bmatrix}
2.5 & 2
\end{bmatrix}
$
	\item % i
$
\begin{bmatrix}
6 & 1
\end{bmatrix}
$
\end{enumerate}

\subsection{Exercise 4}

\begin{enumerate}[a.]
	\item False, the position doesn't matter, size (thus magnitude) does. So, once again: \textit{size matters}. :]
	\item True.
	\item False, you'll end up up in the same place regardless of order.
	\item False, it's the displacement from origin to the point.
\end{enumerate}

\subsection{Exercise 5}

\begin{enumerate}[a.]
	\item % a
$
-\begin{bmatrix}
3 & 7
\end{bmatrix}
=
\begin{bmatrix}
-3 & -7
\end{bmatrix}
$
	\item % b
$
||
\begin{bmatrix}
-12 & 5
\end{bmatrix}
|| = 
\sqrt{(-12)^2+(5)^2}=\sqrt{169}=13
$
	\item % c
$
||
\begin{bmatrix}
8 & -3 & \frac{1}{2}
\end{bmatrix}
|| = 
\sqrt{8^2+(-3)^2+\frac{1}{2}^2} =
\sqrt{64 + 9 + \frac{1}{4}} = \sqrt{73\frac{1}{4}}
$
	\item % d
$
3\begin{bmatrix}
4 & -7 & 0
\end{bmatrix}
=
\begin{bmatrix}
12 & -21 & 0
\end{bmatrix}
$
	\item % e
$
\begin{bmatrix}
4 & 5
\end{bmatrix}/2
=
\begin{bmatrix}
2 & \frac{5}{2}
\end{bmatrix}
$
\end{enumerate}

\end{document}
